Diffusive Nested Sampling: Not Dead Yet

### 2025-11-23 talk/talk.tex
A figure
Integral is over beta
A few slides

\newcommand{\given}{\,|\,}

@@

@@

\begin{frame}
\frametitle{Thermodynamic Integration}
An older technique to calculate $Z$ is based on the following
result:
\begin{align}
\log Z &= \int_0^1 \left<\log L(\theta)\right>_\beta \, d\theta
\end{align}
\pause
where the expectation is taken with respect to `power posteriors'
proportional to $\pi(\theta)L(\theta)^\beta$. \\[0.5em]\pause

@@

>>>>

\newcommand{\given}{\,|\,}
\newcommand{\bx}{\boldsymbol{x}}

@@

\begin{frame}
\frametitle{Bayesian Inference}

\begin{center}
\includegraphics[width=0.7\textwidth]{bayes.pdf}
\end{center}

@@

\begin{frame}
\frametitle{Sampling Strategies}
I divide MCMC (or similar) sampling strategies into two classes.
The first category focuses on {\bf how to move around}:
e.g., Metropolis, Gibbs sampling, Slice Sampling, Hamiltonian MCMC,
Affine Invariant Ensemble Sampler.\\[0.5em]\pause

The second category {\bf modifies the target distribution}.
Examples: Parallel tempering (and other methods based on `annealing'
or `power posteriors'), Nested Sampling, Diffusive Nested Sampling.

\begin{frame}
\frametitle{Thermodynamic Integration}
An older technique to calculate $Z$ is based on the following
result:
\begin{align}
\log Z &= \int_0^1 \left<\log L(\theta)\right>_\beta \, d\beta
\end{align}
\pause
where the expectation is taken with respect to `power posteriors'
proportional to $\pi(\theta)L(\theta)^\beta$. \\[0.5em]\pause

@@

\begin{frame}
\frametitle{Mixture of Two Gaussians}
Taken from John Skilling's 2000s Nested Sampling papers.
$\pi(\bx)$ is Uniform$(-1, 1)$ in 20 dimensions. $L(\bx)$ is a mixture
of a broad and a narrow gaussian:

\begin{align}
L(\bx) &= \prod_{i=1}^{20}
            \frac{\exp\left(-\frac{1}{2}(x_i/v)^2\right)}{v\sqrt{2\pi}}
        + 100 \prod_{i=1}^{20}
            \frac{\exp\left(-\frac{1}{2}(x_i/u)^2\right)}{u\sqrt{2\pi}}.
\end{align}
With $u=0.01$ and $v=0.1$ the narrow peak dominates the posterior (but not
entirely).
